Continous time Markov chain models; \\ extension, simulation and visualisation

### 2014-11-03 chapters/chapter06.tex
first discussion edits

In this concluding chapter we start with highlighting some of the ongoing work, which revolves around using a promising family of priors driven by the Dirichlet process in order to model branch and site-specific parameters of codon substitution models.

Several aspects of the presented work are then put into a wider perspective, with possible extensions discussed in more detail.

Much of the work that constitutes this thesis resolves around relaxing the standard phylogenetic assumptions, in search of more biologically plausible models. 
In that light the work on the time-heterogenous modelling, as presented in Chapter \ref{chap:epoch}, can be viewed as an introductory step towards a broader range of models that tackle different types of time-heterogeneity.
We highlight one possible extension of the epoch model, in which time changes in a non-linear fashion.
We then discuss the possibility to infer both the time and the number of change-points, or transition times, in the epoch model specification via Dirichlet process priors discussed in the first section of this summarizing chapter.

Another substantial portion of the work presented in this thesis is devoted towards development of flexible, easy-to-use software.
%GB: perhaps better to specify more clearly what you mean with phylogenetic and population models?
In that light we talk about the continued effort to support and extend the $\pi$BUSS simulation software \citep{Bielejec2014b}, with an ever-growing array of phylogenetic and population models.

Highly dimensional estimates that result from Bayesian inference of viral spread in both time and space require dedicated software that is capable of producing visualisations that are both visually pleasing and insightful.
We discuss the future directions that the next releases of the SPREAD software \citep{Bielejec2011} will take, ensuring that it keeps providing its users with intuitive and user-friendly interfaces as well as access to a vast bulk of possible visualisations.

Finally we talk about the challenges and opportunities that the future might bring, in the era of the so-called ``genomics plenty'', where vast amounts of molecular data can be sequenced increasingly faster and at lower costs, and the high-performance computing strategies that compete in this race.
\end{remark}

>>>>

In this thesis, we present three pieces of work that revolve around modeling evolution as CTMC process in a phylogenetic context. 
First we demonstrated how we can relax the time-homogeneity assumption in a Bayesian phylogenetic framework leading to the identification of changing sequence and trait evolutionary processes through time.
Next, we introduce a novel simulation architecture that closely matches the model developments in the Bayesian framework we contribute to.
Finally, we complement the Bayesian phylogeographic inference framework with an easy-to-use visualization software to assist the practitioner in interpreting spatiotemporal reconstructions.
In this concluding chapter several aspects of the presented work are put into a wider perspective, and we discuss possible extensions discussed in more detail.

An important part of the work in this thesis is motivated by the need to relax the standard phylogenetic assumptions, in search of more biologically plausible models. 
In that respect the work on the time-heterogenous modeling, as presented in Chapter \ref{chap:epoch}, can be viewed as an introductory step towards a broader range of models that tackle different types of time-heterogeneity.
We highlight one possible extension of the epoch model, in which time changes in a non-linear fashion.
We then discuss the possibility to infer both the time and the number of change-points, or transition times, in the epoch model specification via Dirichlet process priors, on which we elaborate in the second section of this summarizing chapter.

One of the epoch applications focused on the use of codon substitution models to investigate changing selective pressures in HIV evolution within hosts.
Due to their computational burden, the development of codon substitution model extensions in Bayesian phylogenetic inference involving random trees has remained limited.
Motivated by the evolutionary insights dedicated codon substitution models can provide and by the computational speed-ups we achieved in the epoch inference through fine and coarse grain parallelization, we are now starting to explore flexible codon model parameterizations in our framework.  
We highlight some of the ongoing work in which we adopt a promising family of priors driven by the Dirichlet process in order to model among-branch and among-site variation in the codon substitution process.

A substantial portion of the work presented in this thesis is devoted towards development of flexible, easy-to-use software.
%GB: perhaps better to specify more clearly what you mean with phylogenetic and population models?
We therefore elaborate on the continued effort to support and extend the $\pi$BUSS simulation software \citep{Bielejec2014b}, with an ever-growing array of phylogenetic and population models.

Highly dimensional estimates that result from Bayesian inference of viral spread in both time and space require dedicated software capable of producing visualizations that are both visually attractive and insightful.
We discuss the future directions that we would like to take with the next releases of the SPREAD software \citep{Bielejec2011}, ensuring that it continues to provide the users with an intuitive and user-friendly interface as well as access to a variety visualization possibilities.

Finally we talk about the challenges and opportunities that the future might bring, in the era of the so-called `genomics plenty', where vast amounts of molecular data can be sequenced increasingly faster and at lower costs, and the high-performance computing strategies that are required to keep up with this data flood.
\end{remark}
